Solutions to Integer Programming from Quantum Annealing

### 2020-05-25 paper/main.tex
First intro adjustments

@@

Integer linear programming (ILP) is an integer optimization problem subject to inequality constraints and is a commonly tackled problem applicable to situations such as x~\cite{}, y~\cite{}, and z~\cite{}.
In general however, ILP is classically NP-complete, and as a result, heuristic methods are employed~\cite{}.
The NP-hardness of ILP can be understood by realizing that while the solution to an $n$-dimensional linear problem can be obtained in polynomial time, the optimal integer solution in general must be found in the $2^n$ integer solutions which surround the real number solution.
Due to the importance and difficulty, in this work we present a method to obtain the optimal solution to ILP by employing methods of quantum annealing.

@@

Formally, ILP is defined as the optimization of a linear function
{\color{blue}
\begin{align}
    f(x) = &\min\limits_{x}(\sum_i c_i x_i),\\
    \textrm{subject to} \quad & \sum_i A_{ai}x_i +b_a \leq 0 \\
    & x_i \geq 0,\\
    & x_i \in \mathbb{Z}
\end{align}
}
where $i=1, \cdots,  N$ is the number of dependent variables and $a=1, \cdots, M$ the number of constraint equations.

{\color{red} Paragraph on how this is solved classically.
Include strengths and drawbacks of classical solution.}

@@

@@

The inequality constraints can be rewritten as an equality by introducing slack variables $s$ such that the ILP becomes
{\color{blue}
\begin{align}
    f(x) = &\min(\sum_i c_i x_i),\\
    \textrm{subject to} \quad & \sum_i A_{a i}x_i + s_a + b_a = 0,\\
    & s_a \geq 0,\\
    & s_a \in \mathbb{Z},\\
    & x_i \geq 0,\\
    & x_i \in \mathbb{Z}.
\end{align}
}
Since quantum annealing solves the QUBO problem, we extend the proposed algorithm to solve integer quadratic optimization problems such that
{\color{blue}
\begin{align}
    f(x) = \min\limits_{x}(\sum_{ij} x_i d_{ij} x_j + \sum_i c_i x_i)
\end{align}
}
without the introduction of auxiliary qubits.

@@

We map integer variables $z_i$ to qubits under the following transformation~\cite{Chang:2018uoc}
\begin{align}
    z_i = & \sum_{r=0}^{R_i-1} 2^r \psi_{ri}
\label{eq:int_to_bin}
\end{align}
where $\psi_{ri} \in \{0, 1\}$ while the number of qubits used to represent the $i$-th integer variable is allowed to vary with $R_i$.
The transformation can in general be express as a rectangular matrix.
For example, a vector of two integer variables $z_0$ and $z_1$ represented by one and two qubits respectively is given as
\begin{align}
    \begin{pmatrix}
    z_0\\
    z_1
    \end{pmatrix}
    = &
    \begin{pmatrix}
    2^0 & 0 & 0\\
    0 & 2^0 & 2^1
    \end{pmatrix}
    \begin{pmatrix}
    \psi_{00}\\
    \psi_{01}\\
    \psi_{11}
    \end{pmatrix}
    \equiv T^z \begin{pmatrix}
    \psi_{00}\\
    \psi_{01}\\
    \psi_{11}
    \end{pmatrix}
\end{align}
The transformation can be reduced to a tensor product if $R_i$ is a constant for all $i$ such that
\begin{align}
    \mathcal{R} = & \begin{pmatrix} 2^0 & \dots & 2^{R-1}\end{pmatrix},\\
    \mathcal{Z} = & \begin{pmatrix} z_0 & \dots & z_{N-1}\end{pmatrix},\\
    |\mathds{1}| = & |\mathcal{Z}|,\\
    T^z = & \mathds{1}\otimes \mathcal{R}.
\end{align}

While the coefficients of the inequality constraints are not required to be integer valued, the inequalities can be trivially rescaled such that $s_i \in \mathbb{Z}$ given fixed precision coefficients $A_{ij}$ and $b_i$.
As a result, we can restrict both $x_i$ and $s_i$ to sample over only positive integer values as allowed by Eq.~(\ref{eq:int_to_bin}).

@@

The integer quadratic optimization problem can be mapped to a minimization of the quadratic objective function
\begin{align}
    \chi^2 = & \left[\Psi^x_{\mu} T^x_{\mu i}d_{ij} T^x_{j \nu}\Psi^x_\nu + c_i T^x_{i\mu} \Psi^x_\mu \right.
    \nonumber\\
    &\left.+ p (A_{a b} T^x_{b \mu} \Psi^x_{\mu} + T^s_{a \alpha} \Psi^s_\alpha + b_a)^2 \right],\\
    \Psi = & \begin{pmatrix}
    \Psi^x\\
    \Psi^s
    \end{pmatrix},
\end{align}
where $p$ is the strength of the penalty factor which must be set large enough such that the constraints are satisfied under minimization.
The objective function can be represented as a QUBO Hamiltonian
\begin{align}
    E = &
    \begin{pmatrix}
    \Psi^x & \Psi^s
    \end{pmatrix}
    \begin{pmatrix}
    Q_{xx} & Q_{xs}\\
    Q_{sx} & Q_{ss}
    \end{pmatrix}
    \begin{pmatrix}
    \Psi^x\\ \Psi^s
    \end{pmatrix} + pb^2,\\
    \equiv & \Psi^T Q \Psi + C,
    \label{eq:matrix_form}
\end{align}
where
\begin{align}
    Q_{xx} = & {T^{x}}^T \left[ d + p A^T A + p \mathrm{Diag}_{x} \left(A^T b + b^T A\right) \right] T^x \nonumber \\
    &+ \mathrm{Diag}_{\psi^x}(c) T^x,\\
    Q_{xs} = & Q_{sx}^T = p {T^{x}}^T A^T T^s,\\
    Q_{ss} = & p\left[ {T^{s}}^T T^s + \mathrm{Diag}_{\psi^s}\left( {T^{s}}^T b + b^T T^s\right) \right].
\end{align}
The function $\mathrm{Diag}_{z}(v)$ embeds a vector $v$ into a diagonal matrix in subspace $z$, and absorbs the linear contributions of the QUBO into the diagonal elements of the quadratic representation.

%----------------------------------------------------------------------------------------
\subsubsection{Minimum Dominating Set Problem}
\label{sec:methods:mds}
%----------------------------------------------------------------------------------------

@@

%----------------------------------------------------------------------------------------
\subsection{Quantum Annealing}
\label{methods:annealing}
%----------------------------------------------------------------------------------------

%----------------------------------------------------------------------------------------
\subsubsection{Modified annealing schedule}
\label{methods:annealing-schedule}
%----------------------------------------------------------------------------------------

@@

%========================================================================================
\section{CONCLUSION}
%========================================================================================

%========================================================================================
\section{DATA AVAILABILITY}
%========================================================================================

>>>>

\newcommand{\alert}[1]{\textbf{\color{red}{#1}}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

@@

Integer linear programming (ILP) is an integer optimization problem subject to inequality constraints and is a commonly tackled problem applicable to situations such as x~\cite{}, y~\cite{}, and the Minimum Dominating Set Problem (MDS)~\cite{}.
In general however, ILP is classically NP-complete, and as a result, heuristic methods are employed~\cite{}.
The NP-hardness of ILP can be understood by realizing that while the solution to an $n$-dimensional linear problem can be obtained in polynomial time, the optimal integer solution in general must be found in the $2^n$ integer solutions which surround the real number solution.
Standard classical heuristic algorithms employ a greedy scheme which iteratively approximates optimal solutions--starting from a random initial guess, these algorithms apply locally optimal choices at each step.
While greedy algorithms do not guarantee optimal global solutions, they find approximate solutions in polynomial time, which can be utilized in further computations.
Due to the importance and difficulty, in this work we present a method to obtain the optimal solution to ILP by employing methods of quantum annealing.

@@

\alert{[Formulate: Decoherence \& Localization]}
Hardware suffers from...
For this reason one is not guaranteed to find optimal solutions

\alert{[Formulate: Simulations]}
To understand findings and limitations, we mimic the hardware

In section \ref{sec:methods} we propose an algorithm that maps ILP problems to QA hardware (\ref{sec:methods:ilp}) and present an example implementation in case of the Minimial Dominating Set problem (\ref{sec:methods:mds}).
We review the annealing process on the utilized hardware (\ref{sec:methods:annealing}) and describe the computational framework which we used to simulate the annealing process (\ref{sec:methods:simulation}).
Section \ref{sec:results} summarizes our findings and presents...
We conclude this report in \ref{sec:conclusion}.
All algorithms used and data obtained are publicly available (\ref{sec:open-source})

@@

This section presents how a general ILP is mapped to a QUBO.

%----------------------------------------------------------------------------------------
\subsubsection{Integer linear programming}
\label{sec:methods:ilp-def}
%----------------------------------------------------------------------------------------
Formally, ILP is defined as the optimization of a linear function
\begin{align}
    \label{eq:initial-ip-def}
    \vec x_0 = \mathrm{arg} &\min\limits_{x}(\sum_i c_i x_i)
\end{align}
subject to
\begin{align}
    \label{eq:ilp-constraints}
    &\sum_i A_{ai}x_i +b_a \leq 0 \,,\\
    & x_i  \in \mathbbm{N}
\end{align}

where the natural numbers $\mathbbm{N}$ include zero, $i=1, \cdots,  N$ is the number of dependent variables and $a=1, \cdots, M$ the number of constraint equations.

@@

The inequality constraints in eq.~\eqref{eq:ilp-constraints} can be rewritten as an equality by introducing slack variables $s$ such that the ILP becomes
\begin{align}
    \vec x_0 = \mathrm{arg} &\min(\sum_i c_i x_i)\,,
\end{align}
subject to
\begin{align}& \sum_i A_{a i}x_i + s_a + b_a = 0,\\
    & s_a \in \mathbb{N},\\
    & x_i \in \mathbb{N}.
\end{align}
While the coefficients of the inequality constraints are not required to be integer valued, the inequalities can be trivially rescaled such that $s_i \in \mathbbm{N}$ given fixed precision coefficients $A_{ij}$ and $b_i$.
As a result, we can restrict both $x_i$ and $s_i$ to sample over only positive integer values as allowed by Eq.~(\ref{eq:int_to_bin}).

Because quantum annealing is capable of solving quadratic problems, we extend the proposed algorithm to solve integer quadratic optimization problems as well such that
\begin{align}
    \vec x_0 = \mathrm{arg}\min\limits_{x}\left(\sum_{ij} x_i d_{ij} x_j + \sum_i c_i x_i\right)
\end{align}
without the introduction of auxiliary qubits.

@@

We map integer variables $\vec z = \vec x, \vec s$ to qubits under the following transformation~\cite{Chang:2018uoc}
\begin{align}
    z_i = & \sum_{r=0}^{R_i-1} 2^r \psi_{ri}
\label{eq:int_to_bin}
\end{align}
where $\psi_{ri} \in \{0, 1\}$.
The number of qubits used to represent the $i$-th integer variable is allowed to vary with $R_i$.
This transformation can be expressed as a rectangular matrix.
For example, a vector of two integer variables $z_0$ and $z_1$ represented by one and two qubits respectively is given as
\begin{align}
    \begin{pmatrix}
    z_0\\
    z_1
    \end{pmatrix}
    = &
    \begin{pmatrix}
    2^0 & 0 & 0\\
    0 & 2^0 & 2^1
    \end{pmatrix}
    \begin{pmatrix}
    \psi_{00}\\
    \psi_{01}\\
    \psi_{11}
    \end{pmatrix}
    \equiv T^z \begin{pmatrix}
    \psi_{00}\\
    \psi_{01}\\
    \psi_{11}
    \end{pmatrix}
\end{align}
If all variables are represented by the same number of qubits--$R_i$ is a constant for all $i$--then one can express the transformation as tensor product of bit vectors
\begin{align}
    \mathcal{R} = & \begin{pmatrix} 2^0 & \dots & 2^{R-1}\end{pmatrix},\\
    \mathcal{Z} = & \begin{pmatrix} z_0 & \dots & z_{N-1}\end{pmatrix},\\
    |\mathds{1}| = & |\mathcal{Z}|,\\
    T^z = & \mathds{1}\otimes \mathcal{R}.
\end{align}

@@

The integer quadratic optimization problem can be mapped to a minimization of the quadratic objective function
\begin{align}
    \chi^2 = & \left[\Psi^x_{\mu} T^x_{\mu i}d_{ij} T^x_{j \nu}\Psi^x_\nu + c_i T^x_{i\mu} \Psi^x_\mu \right.
    \nonumber\\ \label{eq:qubo-min}
    &\left.+ p (A_{a b} T^x_{b \mu} \Psi^x_{\mu} + T^s_{a \alpha} \Psi^s_\alpha + b_a)^2 \right],\\
    \vec x_0 =&  T^x \vec \Psi^x_0 \, , 
    \qquad (\vec \Psi^x_0, \vec \Psi^s_0) = \mathrm{arg}\min\limits_{\vec{\Psi}^x, \vec \Psi^s}(\chi^2) \,, &
\end{align}
where $p$ is an external parameter representing the strength of the penalty when violating constraints.
The objective function can be represented as a QUBO Hamiltonian
\begin{align}
    E = &
    \begin{pmatrix}
    \Psi^x & \Psi^s
    \end{pmatrix}
    \begin{pmatrix}
    Q_{xx} & Q_{xs}\\
    Q_{sx} & Q_{ss}
    \end{pmatrix}
    \begin{pmatrix}
    \Psi^x\\ \Psi^s
    \end{pmatrix} + pb^2,\\
    \equiv & \Psi^T Q \Psi + C,
    \label{eq:matrix_form}
\end{align}
where
\begin{align}
    Q_{xx} = & {T^{x}}^T \left[ d + p A^T A + p \mathrm{Diag}_{x} \left(A^T b + b^T A\right) \right] T^x \nonumber \\
    &+ \mathrm{Diag}_{\psi^x}(c) T^x,\\
    Q_{xs} = & Q_{sx}^T = p {T^{x}}^T A^T T^s,\\
    Q_{ss} = & p\left[ {T^{s}}^T T^s + \mathrm{Diag}_{\psi^s}\left( {T^{s}}^T b + b^T T^s\right) \right].
\end{align}
The function $\mathrm{Diag}_{z}(v)$ embeds a vector $v$ into a diagonal matrix in subspace $z$, and absorbs the linear contributions of the QUBO into the diagonal elements of the quadratic representation.

%----------------------------------------------------------------------------------------
\subsubsection{Comment on penalty term}
\label{sec:methods:ilp-qubo-comments}
%----------------------------------------------------------------------------------------

The minimal energy solution to eq.~\eqref{eq:initial-ip-def} and eq.~\eqref{eq:qubo-min} are exactly the same if the penalty term is larger than the energy gap of the first excited solution: $p > E_1 - E_0$.
Thus some knowledge of the problem is required.
In principle, it is possible to set the penalty term arbitrarily large, at the cost of problem resolution: large $p$-values increase the highest available energy of the system by multiples of $p$.
After normalization of the QUBO, this corresponds to decreasing the energy gap between the ground state and the first excited state.
Thus, if solvers have finite precision, one must estimate reasonable values for $p$: for large $p$ more solutions fulfill the constraints, while for small $p$, more solutions minimize the objective function.

%----------------------------------------------------------------------------------------
\subsection{Example: Minimum Dominating Set Problem}
\label{sec:methods:mds}
%----------------------------------------------------------------------------------------

%----------------------------------------------------------------------------------------
\subsubsection{Minimum Dominating Set Problem}
\label{sec:methods:mds-def}
%----------------------------------------------------------------------------------------

@@

%----------------------------------------------------------------------------------------
\subsection{Quantum Annealing}
\label{sec:methods:annealing}
%----------------------------------------------------------------------------------------

%----------------------------------------------------------------------------------------
\subsubsection{Modified annealing schedule}
\label{sec:methods:annealing-schedule}
%----------------------------------------------------------------------------------------

@@

%========================================================================================
\section{CONCLUSION}
\label{sec:conclusion}
%========================================================================================

%========================================================================================
\section{DATA AVAILABILITY}
\label{sec:open-source}
%========================================================================================
